---
author: Jeremy Allen
format: html
code-fold: true
---

# Political Violence
This analysis uses data from The Prosecution Project Dataset (Loadenthal et al. 2023) to examine the number of prosecutions for violent crimes across ideological affiliations. The dataset includes information on various criminal methods, ideological affiliations, and prosecution details.

```{r}
#| message: false
#| warning: false

suppressPackageStartupMessages({
    library(tidyverse)
    library(janitor)
    library(RColorBrewer)
})
```

## Prepare the data

Import the data and clean the column names.
```{r}
#| message: false
#| warning: false
df <- read_csv("data/tpp-2025-09-15-general.csv") |> 
    clean_names()
```

How many cases are not linked or motivated politically?
```{r}
#| message: false
#| warning: false
df |>
    filter(criminal_method == "Criminal violation not linked or motivated politically") |> 
    nrow()
```


### Remove cases not politically motivated or unknown, then classify criminal methods as violent or non-violent.
```{r}
#| message: false
#| warning: false

# remove cases not politically motivated or unknown
df <- df |>
    filter(!criminal_method %in% c("Unknown/unspecified/undeveloped", "Criminal violation not linked or motivated politically"))

# classify methods as violent or non-violent
violent <- c(
  "Unarmed assault",
  "Hostage-taking",
  "Armed intimidation/standoff",
  "Vehicle ramming",
  "Chemical or biological weapon deployment",
  "Firearms: civilian",
  "Firearms: military",
  "Explosives",
  "Other weapons"
)

df <- df |> 
    mutate(method_type = if_else(
        criminal_method %in% violent,
        "violent",
        "non-violent"
    ))

```

### Which ideological affiliations are present?
```{r}
#| message: false
#| warning: false
df |> count(ideological_affiliation)
```

### Because there are multiple leftist categories and multiple rightist categories, lets add a column that consolidates leftist types and rightest types.
```{r}
#| message: false
#| warning: false
df <- df |> 
    mutate(
        ideology_simple = case_when(
            str_detect(ideological_affiliation, "Leftist") ~ "Leftist",
            str_detect(ideological_affiliation, "Rightist") ~ "Rightist",
            TRUE ~ ideological_affiliation
        )
    )

df |> count(ideology_simple)
```

### How many prosecutions were for violent vs. non-violent methods, by ideological affiliation?
```{r}
#| message: false
#| warning: false
df |> 
    group_by(ideology_simple, method_type) |> 
    summarise(n = n()) |> 
    pivot_wider(names_from = method_type, values_from = n, values_fill = 0) |>
    select(ideology_simple, violent, `non-violent`) |>
    arrange(desc(violent))
```

## Prosections for violent crimes by ideological affiliation
```{r}
#| message: false
#| warning: false

violent_crimes_by_ideology <- df |> 
    filter(method_type == "violent") |> 
    count(ideology_simple, name = "violent_prosecutions") |> 
    arrange(desc(violent_prosecutions))

ggplot(violent_crimes_by_ideology, aes(x = reorder(ideology_simple, violent_prosecutions), 
                                       y = violent_prosecutions)) +
    geom_col(fill = "black", alpha = 0.8) +
    coord_flip() +
    labs(
        title = "Prosecutions for Violent Crimes by Ideological Affiliation",
        subtitle = "Based on The Prosecution Project Dataset",
        x = "Ideological Affiliation",
        y = "Number of Prosecutions for Violent Crimes",
        caption = "Source: The Prosecution Project, 2023"
    ) +
    theme_minimal() +
    theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 11)
    )
```

## Prosecutions for violent crimes by ideological affiliation by year
```{r}
#| message: false
#| warning: false
#| column: screen-inset

# Extract year from the date column and create the data for the stacked bar chart
df_yearly <- df |>
    # Extract year from date (handling different date formats)
    mutate(year = case_when(
        str_detect(date, "/") ~ as.numeric(str_extract(date, "\\d{4}$")),
        str_detect(date, "-") ~ as.numeric(str_extract(date, "^\\d{4}"))
    )) |>
    # Filter out any rows where year extraction failed
    filter(!is.na(year)) |>
    # Count prosecutions by year and ideology
    count(year, ideology_simple, name = "prosecutions")

# Create a custom color palette that clearly distinguishes ideologies
# Use a manual color scale with intuitive colors
ggplot(df_yearly, aes(x = year, y = prosecutions, fill = ideology_simple)) +
    geom_col() +
    scale_fill_manual(values = c(
        "Rightist" = "#c14a58ff",                    # Red for Rightist
        "Leftist" = "#337fb5ff",                     # Blue for Leftist  
        "Salafi/Jihadist/Islamist" = "#e08738ff",    # Orange
        "Unclear" = "#71a771ff",                     # Green
        "No affiliation/not a factor" = "#999999ff", # Purple
        "Nationalist-separatist" = "#8c564b",      # Brown
        "Other" = "#e377c2"                        # Pink
    )) +
    labs(
        title = "Total Prosecutions by Year, Colored by Ideological Affiliation",
        subtitle = "Based on The Prosecution Project Dataset",
        x = "Year",
        y = "Number of Prosecutions",
        fill = "Ideological Affiliation",
        caption = "Source: The Prosecution Project, 2023"
    ) +
    theme_minimal() +
    theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 11),
        legend.position = "bottom",
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 9)
    ) +
    guides(fill = guide_legend(ncol = 3)) +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 8))
```

## References

Loadenthal, Michael, Lauren Donahoe, Madison Weaver, Sara Godfrey, Kathryn Blowers, et. al. “The Prosecution Project Dataset,” the Prosecution Project, 2023 [dataset]. https://theprosecutionproject.org/

## Code
This descriptive analysis was created by Jeremy Allen using R, tidyverse, janitor, and RColorBrewer packages. The report is built with Quarto. The code is available on [GitHub](https://github.com/jeremy-allen/prosecution.git)
