---
author: Jeremy Allen
format: html
code-fold: true
---

# Political violence in the US

I examine the number of people prosecuted for and number of deaths from violent crimes in the United States. I exclude violent crimes that do not have an ideological motivation and guilty verdict. I use
data[^1] from [The Prosecution Project
Dataset](https://theprosecutionproject.org/overview-project-description/) (Loadenthal et al. 2023). This data covers 1990 to present, though the most recent events are still being processed and may not appear in the data. In this data, each row represents a defendant in a case.

[^1]: See [The Prosecution Project's FAQ](https://theprosecutionproject.org/faq/) where they discuss what data is included, excluded, and why.
    For example, their data is likely an undercount of violence because they compile their data only
    from prosecutions. If there were a violent crime and the perpetrator were killed or died,
    there would be no prosecution and thus no data point in this set for that crime.

After Charlie Kirk was assassinated in September, the US president, Donald Trump, and Elon Musk said
the left is the real problem (Suter 2025; Hutzler and Stoddart 2025). They struggle with numbers, so
this report looks at reliable numbers. The left, right, and others commit violence. However, the
lopsided swell of violence from the right is obvious for any observer interested in knowledge more
than inflammatory politics.

## Prepare the data

Import the data, clean the column names, and remove pending cases and cases without clear guilty
verdict, and filter for only the US.

```{r}
#| message: false
#| warning: false

suppressPackageStartupMessages({
    library(tidyverse)
    library(janitor)
    library(RColorBrewer)
})

jurisdiction_extract <- function(string1, string2) {
  if (is.na(string2)) {
    return(NA)
  }

  string1 <- tolower(string1)
  string2 <- tolower(string2)

  if (string1 == "federal") {
    return("Federal")
  }

  if (!is.character(string2)) {
    stop("string2 must be a character string")
  }

  first_two_words = str_extract(string2, "^\\w+(\\s\\w+)?")

  if (first_two_words == "district of") {
    return("District")
  }

  if (first_two_words == "city of") {
    return("City")
  }

  if (first_two_words == "county of") {
    return("County")
  }

  if (first_two_words == "united states") {
    return("Federal")
  }

  if (
    str_detect(first_two_words, "state") || first_two_words == "commonwealth of"
  ) {
    return("State")
  }

  if (!string1 %in% c("federal", "non-federal")) {
    return("Unknown")
  }

  return("Unknown")

}

```

```{r}
#| message: false
#| warning: false
df <- read_csv(fs::dir_ls("./data", glob = "*.csv")) |> 
    clean_names()

# fix the mix of numeric and character in the # killed and injured columns
df <- df |> 
    mutate(
    number_killed = case_when(
      number_killed %in% c("Multiple", "Unknown") ~ NA_integer_,
      is.na(number_killed) ~ NA_integer_,
      TRUE ~ as.integer(number_killed)
    ),
    number_injured = case_when(
      number_injured %in% c("Multiple", "Unknown") ~ NA_integer_,
      is.na(number_injured) ~ NA_integer_,
      TRUE ~ as.integer(number_injured)
    )
  )

# Add missing case names
df <- df |> 
    mutate(
        name_of_case = str_to_title(name_of_case),
        name_of_case = if_else(
            case_id == "08112021_JMW_WEBB3",
            "State Of Michigan V. Justen Watkins",
            name_of_case
        ),
        name_of_case = if_else(
            case_id == "08112021_TFD_WEBB2",
            "State Of Michigan V. Thomas Denton",
            name_of_case
        ),
        name_of_case = if_else(
            case_id == "08112021_TW_WEBB1",
            "State Of Michigan V. Tristan Webb",
            name_of_case
        ),
        name_of_case = if_else(
            name_of_case == "Data Not Available",
            "Unknown",
            name_of_case
        )
    )

# only US cases
df <- df |> 
    filter(location_country == "United States")

# fix case name typos
df <- df |>
    mutate(
        name_of_case = str_replace_all(name_of_case, "Ofmassachusetts", "Of Massachusetts"),
        name_of_case = str_replace_all(name_of_case, "United Staters", "United States"),
        name_of_case = str_replace_all(name_of_case, "United Staers", "United States"),
        name_of_case = str_replace_all(name_of_case, "United Sates", "United States"),
        name_of_case = str_replace_all(name_of_case, "United Staes", "United States"),
        name_of_case = str_replace_all(name_of_case, "\\bUsa\\b", "United States"),
        name_of_case = str_replace_all(name_of_case, regex("\\bUnited State\\b"), "United States"),
        name_of_case = str_replace_all(name_of_case, regex("\\bUnites States\\b"), "United States"),
        name_of_case = str_remove(name_of_case, regex("^(The )?People Of ")),
        name_of_case = str_replace_all(name_of_case, regex("^The State"), "State"),
        name_of_case = str_replace_all(name_of_case, regex("^Virginia\\b"), "Commonwealth Of Virginia"),
        name_of_case = str_replace_all(name_of_case, regex("^New York\\b"), "State Of New York"),
        name_of_case = str_replace_all(name_of_case, regex("^Ten(n)?essee\\b"), "State Of Tennessee"),
        name_of_case = str_replace_all(name_of_case, regex("^West Virginia\\b"), "State Of West Virginia"),
        name_of_case = str_replace_all(name_of_case, regex("^State V\\. Christopher\\b"), "State Of Nevada V. Christopher"),
        name_of_case = str_replace_all(name_of_case, regex("^State V\\. Finley\\b"), "State Of Nevada V. Finley")
    )

# move appearance number from full names to its own column and clean up
df <- df |> 
  mutate(
    appearance_number = str_extract(full_legal_name, "\\s\\(\\d+\\)$"),
    full_legal_name = str_remove(full_legal_name, "\\s\\(\\d+\\)$"),
    appearance_number = str_remove_all(appearance_number, "[\\(\\)\\s]"),
    appearance_number = as.integer(appearance_number),
    full_legal_name = str_trim(full_legal_name),
    appearance_number = str_trim(appearance_number)
  ) 

# add finer-grained jurisdiction type
df <- df |>
    mutate(
      jurisdiction_level = map2_chr(jurisdiction, name_of_case, jurisdiction_extract)
    )

```

```{r}
no_clear_guilt <- c("Charged but not tried", "Data not available", "Not guilty", "Hung jury/mistrial", "Pending")
num_no_clear_guilt <- df |> 
    filter(verdict %in% no_clear_guilt) |> 
    nrow()

df <- df |> 
    filter(!verdict %in% no_clear_guilt)
```

There are **`r num_no_clear_guilt`** defendants without a clear guilty verdict that I remove from the
analysis.

Are any defendants duplicated for the same charge? For example, being in the data on two different dates as part of the same case? 

```{r}
#| message: false
#| warning: false

duplicates <- df |> 
    group_by(full_legal_name, name_of_case) |> 
    filter(n() > 1) |> 
    ungroup()

df <- df |>
    filter(!case_id %in% c("09262001_PMC", "02142024_KC"))

# Remove incorrect leftist classification for Cody Michael Tarner (should be rightist)
df <- df |>
    filter(!(case_id == "09082020_CMT" & ideological_affiliation == "Leftist: government-focused"))

# Remove incorrect nationalist-separatist classification for Jonathan L. Xie (should be Salafi/Jihadist/Islamist)
df <- df |>
    filter(!(case_id == "09152020_JLX" & ideological_affiliation == "Nationalist-separatist"))

# Remove incorrect "Carried through" classification for Cole James Bridges (should be Attempted)
df <- df |>
    filter(!(case_id == "02032021_CJB" & completion_of_crime == "Carried through"))

# Remove complaint row for John Malcolm Bareswill (keep indictment only)
df <- df |>
    filter(!(case_id == "06112020_JMB" & date_descriptor == "Complaint"))

# Remove incorrect "Not guilty" row for Daniel Delbert Dorson (keep guilty plea row)
df <- df |>
    filter(!(case_id == "12162020_DDD_MILLER4"))

# Remove initial indictment for Adis Medunjanin (keep superseding indictment from July 2010)
df <- df |>
    filter(!(case_id == "02252010_AM_AHMEDZAY2"))

# Remove initial indictment for Thomas Heard (keep later indictment with full charges)
df <- df |>
    filter(!(case_id == "03092019_TH"))

# Remove "Charged but not tried" row for Anthony Comello (keep guilty plea row)
df <- df |>
    filter(!(case_id == "04282019_AC"))

# Remove arrest row for Jason Brown (keep indictment row)
df <- df |>
    filter(!(case_id == "11142019_JB" & date_descriptor == "Arrest/arraignment"))

# Remove complaint row for Marian Hudak (keep indictment row)
df <- df |>
    filter(!(case_id == "06222023_MH" & date_descriptor == "Complaint"))

# Remove indictment row for Rowan McManigal (keep sentencing row)
df <- df |>
    filter(!(case_id == "04042022_RM" & date_descriptor == "Indictment"))

# Remove initial indictment for Walter Edmund Bond (keep later consolidated indictment)
df <- df |>
    filter(!(case_id == "07272010_WEB"))

# Remove indictment row for Michael David Fox (keep plea row)
df <- df |>
    filter(!(case_id == "09202023_MDF" & date_descriptor == "Indictment"))

# Remove arrest row for Omar Alkattoul (keep indictment row)
df <- df |>
    filter(!(case_id == "11102022_OA" & date_descriptor == "Arrest/arraignment"))

```

There are eight cases where defendants appear in the data multiple times for the same charge but at different stages of prosecution (e.g., indictment, arrest, plea, sentencing). I remove eight rows to avoid double counting these defendants, keeping the row that represents the most complete or final stage of their case. This leaves us with **`r nrow(df)`** defendants.

How many defendants are not linked or motivated politically?

```{r}
#| message: false
#| warning: false
total_defendants <- nrow(df)

defendants_not_political <- df |>
    filter(criminal_method == "Criminal violation not linked or motivated politically") |> 
    nrow()
```

There are **`r defendants_not_political`** defendants not politically motivated out of a total of
**`r total_defendants` defendants**, which is **`r round((defendants_not_political / total_defendants) * 100, 2)`%**.
So, I exclude these non-politically motivated defendants going forward.

```{r}
#| message: false
#| warning: false

# remove defendants not politically motivated or unknown
df <- df |>
    filter(!criminal_method %in% c("Unknown/unspecified/undeveloped", "Criminal violation not linked or motivated politically"))

# parse dates
df <- df |> 
    mutate(
        date = str_trim(date),
        date = lubridate::mdy(date)
    ) 

```

Were all of these crimes fully carried through?

```{r}
#| message: false
#| warning: false
df |> 
    count(completion_of_crime) |> 
    knitr::kable(
        col.names = c("Case Outcome", "Number of Prosecutions"),
        caption = "Table 1: Case Outcomes"
    )

num_threats_and_unknown <- df |> 
    filter(completion_of_crime %in% c("Threat", "Unknown")) |> 
    nrow()
```

I am more interested in harm that was attempted, planned, or occurred, so I will filter out the others (`r num_threats_and_unknown` defendants). That leaves us with **`r nrow(df) - num_threats_and_unknown` defendants.**

```{r}
#| message: false
#| warning: false
df <- df |> 
    filter(!completion_of_crime %in% c("Threat", "Unknown"))
```

Which dates are represented in the data?

```{r}
#| message: false
#| warning: true

# Warn if minimum date is earlier than 1900 or latest date is later than today
date_range <- range(df$date, na.rm = TRUE)
if (date_range[1] < as.Date("1900-01-01") || date_range[2] > Sys.Date()) {
    warning("Date range is outside expected bounds.")
}


df |>
    summarise(
        `Earliest Date` = min(date, na.rm = TRUE),
        `Latest Date` = max(date, na.rm = TRUE)
    ) |> 
    knitr::kable(caption = "Table 2: Date Range of the Data")
```

### Which ideological affiliations are represented in the data?

```{r}
#| message: false
#| warning: false
df |>
    count(ideological_affiliation) |> 
    knitr::kable(
        col.names = c("Ideological Affiliation", "Number of Prosecutions"),
        caption = "Table 3: Number of Prosecutions by Ideological Affiliation"
    )
```

### Simplify ideologies

Because there are multiple leftist categories and multiple rightist categories, I add a column
that consolidates leftist types and rightest types. This hides important variation **within** ideologies but allows to more easily asses political statements about left vs. right violence.

```{r}
#| message: false
#| warning: false
df <- df |> 
    mutate(
        ideology_simple = case_when(
            str_detect(ideological_affiliation, "Leftist") ~ "Leftist",
            str_detect(ideological_affiliation, "Rightist") ~ "Rightist",
            TRUE ~ ideological_affiliation
        )
    )

df |>
    count(ideology_simple, sort = TRUE) |> 
    knitr::kable(
        col.names = c("Ideological Affiliation", "Number of Prosecutions"),
        caption = "Table 4: Number of Prosecutions by Simplified Ideological Affiliation"
    )
```

# Focus on Violence

### How many defendants were used violent vs. non-violent methods, by ideological affiliation?

Then I filter for only those to make our visualizations.

```{r}
#| message: false
#| warning: false

# classify methods as violent or non-violent
violent <- c(
  "Unarmed assault",
  "Hostage-taking",
  "Armed intimidation/standoff",
  "Vehicle ramming",
  "Chemical or biological weapon deployment",
  "Firearms: civilian",
  "Firearms: military",
  "Explosives",
  "Other weapons"
)

df <- df |> 
    mutate(method_type = if_else(
        criminal_method %in% violent,
        "violent",
        "non-violent"
    ))

df |> 
    group_by(ideology_simple, method_type) |> 
    summarise(n = n()) |> 
    pivot_wider(names_from = method_type, values_from = n, values_fill = 0) |>
    select(ideology_simple, violent, `non-violent`) |>
    arrange(desc(violent)) |> 
    knitr::kable(
        col.names = c("Ideological Affiliation", "Violent Prosecutions", "Non-Violent Prosecutions"),
        caption = "Table 5: Number of Violent vs. Non-Violent Prosecutions by Ideological Affiliation"
    )

```

## Defendants prosecuted for violent crimes in the US by ideological affiliation

```{r}
#| message: false
#| warning: false
#| fig-width: 11

violent_crimes <- df |> 
    filter(
        method_type == "violent",
        ideology_simple != "No affiliation/not a factor"
        )
    
violent_crimes_by_ideology <- violent_crimes |> 
    count(ideology_simple, name = "violent_prosecutions") |> 
    arrange(desc(violent_prosecutions))

ggplot(violent_crimes_by_ideology, aes(x = reorder(ideology_simple, violent_prosecutions), 
                                       y = violent_prosecutions)) +
    geom_col(fill = "black", alpha = 0.8) +
    coord_flip() +
    labs(
        title = "Defendants Prosecuted for Violent Crimes in the US",
        subtitle = "By Ideological Affiliation of the Perpetrator",
        x = "Their Ideology",
        y = "Number of Defendants",
        caption = "Source: The Prosecution Project"
    ) +
    theme_minimal() +
    theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 11)
    )
```

## Defendants prosecuted for violent crimes in the US by ideological affiliation by year

```{r}
#| message: false
#| warning: false
#| fig-width: 13
#| column: page

df_yearly <- violent_crimes |>
    mutate(
        year = lubridate::year(date),
        year = as.integer(year)
    ) |>
    filter(!is.na(year)) |>
    count(year, ideology_simple, name = "prosecutions")

presidents <- tibble(
  year = c(1993L, 2001L, 2009L, 2017L, 2021L, 2025L),
  president = c("Clinton (D)", "Bush (R)", "Obama (D)", "Trump (R)", "Biden (D)", "Trump (R)")
)

ggplot() +
    geom_vline(data = presidents, aes(xintercept = year), 
               color = "black", linetype = "dashed", alpha = 0.7) +
    geom_text(data = presidents, aes(x = year, y = 90, label = president), 
              vjust = 1.5, hjust = 0, angle = 90, size = 3, color = "black") +
    geom_col(data = df_yearly, aes(x = year, y = prosecutions, fill = ideology_simple)) +
    scale_fill_manual(values = c(
        "Rightist" = "#c14a58ff",                    # Red for Rightist
        "Leftist" = "#337fb5ff",                     # Blue for Leftist  
        "Salafi/Jihadist/Islamist" = "#e08738ff",    # Orange
        "Unclear" = "#71a771ff",                     # Green
        "Nationalist-separatist" = "#8c564b",      # Brown
        "Other" = "#e377c2"                        # Pink
    )) +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
    labs(
        title = "Defendants Prosecuted for Violent Crimes in the US by Year",
        subtitle = "By Ideological Affiliation of the Perpetrator",
        x = "Year",
        y = "Number of Defendants",
        fill = "Their Ideology",
        caption = "Source: The Prosecution Project"
    ) +
    guides(fill = guide_legend(ncol = 6)) +
    theme_minimal() +
    theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 11),
        panel.grid.minor = element_blank(),
        legend.position = "bottom",
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 9)
    )  

```

## Deaths

The data is structured by prosecution. A single defendant can be prosecuted in multiple jurisdictions for the same incident. For example, a shooter can appear in a federal indictment and a state indictment for the same shooting. If I am not careful I could overcount deaths from these crimes. 

How often does this happen? How much would we overcount if we simply summed the number killed from all rows in the data?
```{r}
# were any individuals charged at multiple jurisdictional levels?
dupes <- violent_crimes |>
  group_by(full_legal_name) |>
  arrange(date) |>
  summarize(
    n_jurisdiction_levels = n_distinct(jurisdiction_level),
    jurisdictions = str_c(jurisdiction_level, collapse = "; "),
    case_ids = str_c(case_id, collapse = "; "),
    date_types = str_c(date_descriptor, collapse = "; "),
    state_locations = str_c(location_state, collapse = "; "),
    n_killed_each = str_c(number_killed, collapse = "; "),
    ideology = str_c(ideology_simple, collapse = "; ")
  ) |>
  filter(n_jurisdiction_levels > 1) |>
  arrange((full_legal_name)) |>
  select(-n_jurisdiction_levels)

# how much overcount?
# extract digits from n_killed_each column where there are multiple numbers that are the same
dupe_counts <- dupes |>
  mutate(
    n_killed_each_list = str_extract_all(n_killed_each, "\\d+"),
    n_killed_each_list = map(n_killed_each_list, as.integer),
    total_killed_if_summed = map_int(n_killed_each_list, sum),
    are_equal = map_lgl(n_killed_each_list, ~ length(unique(.x)) == 1)
  ) |>
  filter(are_equal) |>  # just the ones where the numbers are the same
  mutate(actual_killed = map_int(n_killed_each_list, ~ ifelse(length(.x) > 0, .x[1], 0))) |>
  select(-n_killed_each_list, -are_equal)

dupes |> 
    knitr::kable(
        col.names = c("Defendant Name", "Jurisdictions", "Case IDs", "Date Types", "States", "# Killed Each", "Ideology"),
        caption = "Table 6: Defendants Charged at Multiple Jurisdictional Levels"
    )
```

Out of our **`r nrow(violent_crimes)`** violent crime prosecutions, there are **`r nrow(dupes)`** defendants who were charged at multiple jurisdiction levels. If I simply summed the number killed from every row, I would mistakenly double count some deaths.  For these 11 defendants, I would get **`r sum(dupe_counts$total_killed_if_summed, na.rm = TRUE)`** deaths instead of the actual **`r sum(dupe_counts$actual_killed, na.rm = TRUE)`** deaths. This would be  an overcount of **`r sum(dupe_counts$total_killed_if_summed, na.rm = TRUE) - sum(dupe_counts$actual_killed, na.rm = TRUE)`** deaths.

Now, let's look at the number of deaths by ideological affiliation, and avoid double counting from cases where a defendant was prosecuted twice.

```{r}
#| message: false
#| warning: false
#| fig-width: 11

# Don't double count deaths in cases where more than one defendant is tried for the same case
df_deaths <- violent_crimes |>
  mutate(
    # Extract the incident date from Case ID (first part before underscore)  
    incident_date = str_extract(case_id, "^[^_]+"),
    # Parse the date correctly
    incident_date_formatted = case_when(
      str_length(incident_date) == 8 & str_detect(incident_date, "^\\d{8}$") ~ 
        paste0(str_sub(incident_date, 1, 2), "/", 
               str_sub(incident_date, 3, 4), "/", 
               str_sub(incident_date, 5, 8)),
      TRUE ~ incident_date
    )
  ) |>
  group_by(incident_date, incident_date_formatted) |>
  summarise(
    unique_incidents = 1,  
    related_cases = n(),   
    total_individuals = n_distinct(full_legal_name), 
    total_deaths = first(number_killed, na_rm = TRUE),  
    case_names = paste(unique(name_of_case), collapse = "; "),
    ideology_simple = first(ideology_simple, na_rm = TRUE),
    .groups = "drop"
  ) |>
  group_by(ideology_simple) |> 
  summarise(total_deaths = sum(total_deaths, na.rm = TRUE)) |> 
  arrange(desc(total_deaths))

plot1 <- ggplot(df_deaths, aes(x = reorder(ideology_simple, total_deaths), 
                                       y = total_deaths)) +
    geom_col(fill = "black", alpha = 0.8) +
    coord_flip() +
    labs(
        title = "Deaths Reported in Prosecutions in the US",
        subtitle = "By Ideological Affiliation of the Perpetrator",
        x = "Perpetrator Ideology",
        y = "Deaths",
        caption = "Source: The Prosecution Project"
    ) +
    theme_minimal() +
    theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 11)
    )

plot1

```
## Notable Trends

An obvious spike for defendants prosecuted during the George Floyd protests in 2020.

```{r}
#| message: false
#| warning: false

# just leftist
yearly_left <- df_yearly |> 
    filter(ideology_simple == "Leftist")

ggplot() +
    geom_vline(data = presidents, aes(xintercept = year), 
               color = "black", linetype = "dashed", alpha = 0.7) +
    geom_text(data = presidents, aes(x = year, y = 38, label = president), 
              vjust = 1.5, hjust = 0, angle = 90, size = 3, color = "black") +
    geom_col(data = yearly_left, aes(x = year, y = prosecutions, fill = ideology_simple)) +
    scale_fill_manual(values = c(
        "Rightist" = "#c14a58ff",                    # Red for Rightist
        "Leftist" = "#337fb5ff",                     # Blue for Leftist  
        "Salafi/Jihadist/Islamist" = "#e08738ff",    # Orange
        "Unclear" = "#71a771ff",                     # Green
        "Nationalist-separatist" = "#8c564b",      # Brown
        "Other" = "#e377c2"                        # Pink
    )) +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
    labs(
        title = "Leftist Defendants Prosecuted for Violent Crimes in the US by Year",
        subtitle = "",
        x = "Year",
        y = "Number of Defendants",
        fill = "Their Ideology",
        caption = "Source: The Prosecution Project"
    ) +
    guides(fill = guide_legend(ncol = 6)) +
    theme_minimal() +
    theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 11),
        panel.grid.minor = element_blank(),
        legend.position = "bottom",
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 9)
    )  

```

And an obvious spike for defendants prosecuted following the January 6 Capital riot in 2021.

```{r}
#| message: false
#| warning: false

# just leftist
yearly_right <- df_yearly |> 
    filter(ideology_simple == "Rightist")

ggplot() +
    geom_vline(data = presidents, aes(xintercept = year), 
               color = "black", linetype = "dashed", alpha = 0.7) +
    geom_text(data = presidents, aes(x = year, y = 80, label = president), 
              vjust = 1.5, hjust = 0, angle = 90, size = 3, color = "black") +
    geom_col(data = yearly_right, aes(x = year, y = prosecutions, fill = ideology_simple)) +
    scale_fill_manual(values = c(
        "Rightist" = "#c14a58ff",                    # Red for Rightist
        "Leftist" = "#337fb5ff",                     # Blue for Leftist  
        "Salafi/Jihadist/Islamist" = "#e08738ff",    # Orange
        "Unclear" = "#71a771ff",                     # Green
        "Nationalist-separatist" = "#8c564b",      # Brown
        "Other" = "#e377c2"                        # Pink
    )) +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
    labs(
        title = "Rightist Defendants Prosecuted for Violent Crimes in the US by Year",
        subtitle = "",
        x = "Year",
        y = "Number of Defendants",
        fill = "Their Ideology",
        caption = "Source: The Prosecution Project"
    ) +
    guides(fill = guide_legend(ncol = 6)) +
    theme_minimal() +
    theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 11),
        panel.grid.minor = element_blank(),
        legend.position = "bottom",
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 9)
    )  
```
Perhaps the most notable trend is the rise in the number of defendants whose ideology is unclear.

```{r}
#| message: false
#| warning: false

yearly_unclear <- df_yearly |> 
    filter(ideology_simple == "Unclear")

ggplot() +
    geom_vline(data = presidents, aes(xintercept = year), 
               color = "black", linetype = "dashed", alpha = 0.7) +
    geom_text(data = presidents, aes(x = year, y = 12, label = president), 
              vjust = 1.5, hjust = 0, angle = 90, size = 3, color = "black") +
    geom_col(data = yearly_unclear, aes(x = year, y = prosecutions, fill = ideology_simple)) +
    scale_fill_manual(values = c(
        "Rightist" = "#c14a58ff",                    # Red for Rightist
        "Leftist" = "#337fb5ff",                     # Blue for Leftist  
        "Salafi/Jihadist/Islamist" = "#e08738ff",    # Orange
        "Unclear" = "#71a771ff",                     # Green
        "Nationalist-separatist" = "#8c564b",      # Brown
        "Other" = "#e377c2"                        # Pink
    )) +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
    labs(
        title = "Defendants Prosecuted for Violent Crimes in the US by Year",
        subtitle = "Whose ideology is unclear",
        x = "Year",
        y = "Number of Defendants",
        fill = "Their Ideology",
        caption = "Source: The Prosecution Project"
    ) +
    guides(fill = guide_legend(ncol = 6)) +
    theme_minimal() +
    theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 11),
        panel.grid.minor = element_blank(),
        legend.position = "bottom",
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 9)
    )
```

## Unclear Ideological Affiliation

Some defendants in the dataset do not have a clear ideological affiliation. These cases represent violent acts where the motivation is ambiguous or not clearly tied to a specific political ideology. This section examines what methods these defendants used and what kinds of targets they attacked.

```{r}
#| message: false
#| warning: false

unclear_ideology <- violent_crimes |>
    filter(ideology_simple == "Unclear")

# Criminal methods for unclear ideology
method_counts <- unclear_ideology |>
    count(criminal_method, sort = TRUE) |>
    filter(!is.na(criminal_method)) |>
    slice_head(n = 15)

ggplot(method_counts, aes(x = reorder(criminal_method, n), y = n)) +
    geom_col(fill = "black", alpha = 0.8) +
    coord_flip() +
    labs(
        title = "Criminal Methods Used by Defendants",
        subtitle = "With Unclear Ideological Affiliation",
        x = NULL,
        y = "Number of Defendants",
        caption = "Source: The Prosecution Project"
    ) +
    theme_minimal() +
    theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 11)
    )
```

```{r}
#| message: false
#| warning: false

# Ideological targets for unclear ideology
target_counts <- unclear_ideology |>
    count(`ideological_target`, sort = TRUE) |>
    filter(!is.na(ideological_target)) |>
    slice_head(n = 15)

ggplot(target_counts, aes(x = reorder(ideological_target, n), y = n)) +
    geom_col(fill = "black", alpha = 0.8) +
    coord_flip() +
    labs(
        title = "Targets Attacked by Defendants",
        subtitle = "With Unclear Ideological Affiliation",
        x = NULL,
        y = "Number of Defendants",
        caption = "Source: The Prosecution Project"
    ) +
    theme_minimal() +
    theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 11)
    )
```

## References

Hutzler, Alexandra, and Michelle Stoddart. 2025. “Trump Doubles down on Blaming ‘radical Left’ after
Vow to Go after Political Violence.” ABC News, September 12.
https://abcnews.go.com/Politics/trump-doubles-blaming-radical-left-after-vow-after/story?id=125509965.

Loadenthal, Michael, Lauren Donahoe, Madison Weaver, Sara Godfrey, Kathryn Blowers, et. al. 2023.
“The Prosecution Project Dataset,” the Prosecution Project, 2023 \[dataset\].
https://theprosecutionproject.org/

Suter, Tara. 2025. “Elon Musk: ‘The Left Is the Party of Murder.’” Text. The Hill, September 14.
https://thehill.com/policy/technology/5502535-elon-musk-charlie-kirk-death/.

## Code

This descriptive analysis was created by Jeremy Allen using R, tidyverse, janitor, and RColorBrewer
packages. The report is built with Quarto. The code is available on
[GitHub](https://github.com/jeremy-allen/prosecution.git).