---
author: Jeremy Allen
format: html
code-fold: true
---

# Political violence in the US

I examine the number of people prosecuted and number of deaths from violent crimes in the United States. Not all
violent crimes, just violent crimes that have an ideological motivation and guilty verdict. I use
data[^1] from [The Prosecution Project
Dataset](https://theprosecutionproject.org/overview-project-description/) (Loadenthal et al. 2023). In this data, each row is a defendant in a case.

[^1]: See [The Prosecution Project's FAQ](https://theprosecutionproject.org/faq/) where they discuss what data is included, excluded, and why.
    For example, their data is likely an undercount of violence because they compile their data only
    from prosecutions. So, if there were a violent incident and the perpetrator were killed onsite,
    there would be no prosecution and no data point in this set.

After Charlie Kirk was assassinated in September, the US president, Donald Trump, and Elon Musk said
the left is the real problem (Suter 2025; Hutzler and Stoddart 2025). They struggle with numbers, so
this report looks at reliable numbers. The left, right, and others commit violence. However, the
lopsided swell of violence from the right is obvious for any observer interested in knowledge more
than inflammatory politics.

## Prepare the data

Import the data, clean the column names, and remove pending cases and cases without clear guilty
verdict, and filter for only the US.

```{r}
#| message: false
#| warning: false

suppressPackageStartupMessages({
    library(tidyverse)
    library(janitor)
    library(RColorBrewer)
})
```

```{r}
#| message: false
#| warning: false
df <- read_csv("data/tpp-2025-09-15-general.csv") |> 
    clean_names()

# fix the mix of numeric and character in the # killed column
df <- df |> 
    mutate(
    number_killed = case_when(
      number_killed %in% c("Multiple", "Unknown") ~ NA_real_,  # Convert text to NA
      is.na(number_killed) ~ NA_real_,  # Keep existing NAs
      TRUE ~ number_killed  # Keep numeric values as they are
    )
  )

no_clear_guilt <- c("Charged but not tried", "Data not available", "Not guilty", "Hung jury/mistrial", "Pending")
num_no_clear_guilt <- df |> 
    filter(verdict %in% no_clear_guilt) |> 
    nrow()

df <- df |> 
    filter(!verdict %in% no_clear_guilt)

df <- df |> 
    filter(location_country == "United States")
```

There are **`r num_no_clear_guilt`** defendants without a clear guilty verdict that I remove from the
analysis.

How many defendants are not linked or motivated politically?

```{r}
#| message: false
#| warning: false
total_defendants <- nrow(df)

defendants_not_political <- df |>
    filter(criminal_method == "Criminal violation not linked or motivated politically") |> 
    nrow()
```

There are **`r defendants_not_political`** defendants not politically motivated out of a total of
**`r total_defendants` defendants**, which is **`r round((defendants_not_political / total_defendants) * 100, 2)`%**.
So, I exclude these non-politically motivated defendants going forward.

```{r}
#| message: false
#| warning: false

# remove defendants not politically motivated or unknown
df <- df |>
    filter(!criminal_method %in% c("Unknown/unspecified/undeveloped", "Criminal violation not linked or motivated politically"))

# parse dates
df <- df |> 
    mutate(
        date = str_trim(date),
        date = lubridate::mdy(date)
    ) 

```

Were all of these crimes fully carried through?

```{r}
#| message: false
#| warning: false
df |> 
    count(completion_of_crime) |> 
    knitr::kable(
        col.names = c("Case Outcome", "Number of Prosecutions"),
        caption = "Table 1: Case Outcomes"
    )

num_threats_and_unknown <- df |> 
    filter(completion_of_crime %in% c("Threat", "Unknown")) |> 
    nrow()
```

I am more interested in harm that was attempted, planned, or occurred, so I will filter out the others (`r num_threats_and_unknown` defendants). That leaves us with **`r nrow(df) - num_threats_and_unknown` defendants.**

```{r}
#| message: false
#| warning: false
df <- df |> 
    filter(!completion_of_crime %in% c("Threat", "Unknown"))
```

Which dates are represented in the data?

```{r}
#| message: false
#| warning: true

# Warn if minimum date is earlier than 1900 or latest date is later than today
date_range <- range(df$date, na.rm = TRUE)
if (date_range[1] < as.Date("1900-01-01") || date_range[2] > Sys.Date()) {
    warning("Date range is outside expected bounds.")
}


df |>
    summarise(
        `Earliest Date` = min(date, na.rm = TRUE),
        `Latest Date` = max(date, na.rm = TRUE)
    ) |> 
    knitr::kable(caption = "Table 2: Date Range of the Data")
```

### Which ideological affiliations are represented in the data?

```{r}
#| message: false
#| warning: false
df |>
    count(ideological_affiliation) |> 
    knitr::kable(
        col.names = c("Ideological Affiliation", "Number of Prosecutions"),
        caption = "Table 3: Number of Prosecutions by Ideological Affiliation"
    )
```

### Simplify ideologies

Because there are multiple leftist categories and multiple rightist categories, I add a column
that consolidates leftist types and rightest types. This hides important variation **within** ideologies but allows to more easily asses political statements about left vs. right violence.

```{r}
#| message: false
#| warning: false
df <- df |> 
    mutate(
        ideology_simple = case_when(
            str_detect(ideological_affiliation, "Leftist") ~ "Leftist",
            str_detect(ideological_affiliation, "Rightist") ~ "Rightist",
            TRUE ~ ideological_affiliation
        )
    )

df |>
    count(ideology_simple, sort = TRUE) |> 
    knitr::kable(
        col.names = c("Ideological Affiliation", "Number of Prosecutions"),
        caption = "Table 4: Number of Prosecutions by Simplified Ideological Affiliation"
    )
```

# Focus on Violence

### How many defendants were used violent vs. non-violent methods, by ideological affiliation?

Then I filter for only those to make our visualizations.

```{r}
#| message: false
#| warning: false

# classify methods as violent or non-violent
violent <- c(
  "Unarmed assault",
  "Hostage-taking",
  "Armed intimidation/standoff",
  "Vehicle ramming",
  "Chemical or biological weapon deployment",
  "Firearms: civilian",
  "Firearms: military",
  "Explosives",
  "Other weapons"
)

df <- df |> 
    mutate(method_type = if_else(
        criminal_method %in% violent,
        "violent",
        "non-violent"
    ))

df |> 
    group_by(ideology_simple, method_type) |> 
    summarise(n = n()) |> 
    pivot_wider(names_from = method_type, values_from = n, values_fill = 0) |>
    select(ideology_simple, violent, `non-violent`) |>
    arrange(desc(violent)) |> 
    knitr::kable(
        col.names = c("Ideological Affiliation", "Violent Prosecutions", "Non-Violent Prosecutions"),
        caption = "Table 5: Number of Violent vs. Non-Violent Prosecutions by Ideological Affiliation"
    )

```

## Defendants prosecuted for violent crimes in the US by ideological affiliation

```{r}
#| message: false
#| warning: false
#| fig-width: 11

violent_crimes <- df |> 
    filter(
        method_type == "violent",
        ideology_simple != "No affiliation/not a factor"
        )
    
violent_crimes_by_ideology <- violent_crimes |> 
    count(ideology_simple, name = "violent_prosecutions") |> 
    arrange(desc(violent_prosecutions))

ggplot(violent_crimes_by_ideology, aes(x = reorder(ideology_simple, violent_prosecutions), 
                                       y = violent_prosecutions)) +
    geom_col(fill = "black", alpha = 0.8) +
    coord_flip() +
    labs(
        title = "Defendants Prosecuted for Violent Crimes in the US",
        subtitle = "By Ideological Affiliation of the Perpetrator",
        x = "Their Ideology",
        y = "Number of Defendants",
        caption = "Source: The Prosecution Project"
    ) +
    theme_minimal() +
    theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 11)
    )
```

## Defendants prosecuted for violent crimes in the US by ideological affiliation by year

```{r}
#| message: false
#| warning: false
#| fig-width: 13
#| column: page

df_yearly <- violent_crimes |>
    mutate(
        year = lubridate::year(date),
        year = as.integer(year)
    ) |>
    filter(!is.na(year)) |>
    count(year, ideology_simple, name = "prosecutions")

ggplot(df_yearly, aes(x = year, y = prosecutions, fill = ideology_simple)) +
    geom_col() +
    scale_fill_manual(values = c(
        "Rightist" = "#c14a58ff",                    # Red for Rightist
        "Leftist" = "#337fb5ff",                     # Blue for Leftist  
        "Salafi/Jihadist/Islamist" = "#e08738ff",    # Orange
        "Unclear" = "#71a771ff",                     # Green
        "Nationalist-separatist" = "#8c564b",      # Brown
        "Other" = "#e377c2"                        # Pink
    )) +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
    labs(
        title = "Defendants Prosecuted for Violent Crimes in the US by Year",
        subtitle = "By Ideological Affiliation of the Perpetrator",
        x = "Year",
        y = "Number of Defendants",
        fill = "Their Ideology",
        caption = "Source: The Prosecution Project"
    ) +
    guides(fill = guide_legend(ncol = 6)) +
    theme_minimal() +
    theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 11),
        legend.position = "bottom",
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 9)
    )
```

## Deaths

```{r}
#| message: false
#| warning: false
#| fig-width: 11

# Don't double count deaths in cases where more than one defendant is tried for the same case
df_deaths <- violent_crimes |>
  mutate(
    # Extract the incident date from Case ID (first part before underscore)  
    incident_date = str_extract(case_id, "^[^_]+"),
    # Parse the date correctly
    incident_date_formatted = case_when(
      str_length(incident_date) == 8 & str_detect(incident_date, "^\\d{8}$") ~ 
        paste0(str_sub(incident_date, 1, 2), "/", 
               str_sub(incident_date, 3, 4), "/", 
               str_sub(incident_date, 5, 8)),
      TRUE ~ incident_date
    )
  ) |>
  group_by(incident_date, incident_date_formatted) |>
  summarise(
    unique_incidents = 1,  
    related_cases = n(),   
    total_individuals = n_distinct(full_legal_name), 
    total_deaths = first(number_killed, na_rm = TRUE),  
    case_names = paste(unique(name_of_case), collapse = "; "),
    ideology_simple = first(ideology_simple, na_rm = TRUE),
    .groups = "drop"
  ) |>
  group_by(ideology_simple) |> 
  summarise(total_deaths = sum(total_deaths, na.rm = TRUE)) |> 
  arrange(desc(total_deaths))

plot1 <- ggplot(df_deaths, aes(x = reorder(ideology_simple, total_deaths), 
                                       y = total_deaths)) +
    geom_col(fill = "black", alpha = 0.8) +
    coord_flip() +
    labs(
        title = "Who Causes More Death by Violent Crime in the US?",
        subtitle = "By Ideological Affiliation of the Perpetrator",
        x = "Perpetrator Ideology",
        y = "Deaths",
        caption = "Source: The Prosecution Project"
    ) +
    theme_minimal() +
    theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 11)
    )

plot1

```


## References

Hutzler, Alexandra, and Michelle Stoddart. 2025. “Trump Doubles down on Blaming ‘radical Left’ after
Vow to Go after Political Violence.” ABC News, September 12.
https://abcnews.go.com/Politics/trump-doubles-blaming-radical-left-after-vow-after/story?id=125509965.

Loadenthal, Michael, Lauren Donahoe, Madison Weaver, Sara Godfrey, Kathryn Blowers, et. al. 2023.
“The Prosecution Project Dataset,” the Prosecution Project, 2023 \[dataset\].
https://theprosecutionproject.org/

Suter, Tara. 2025. “Elon Musk: ‘The Left Is the Party of Murder.’” Text. The Hill, September 14.
https://thehill.com/policy/technology/5502535-elon-musk-charlie-kirk-death/.

## Code

This descriptive analysis was created by Jeremy Allen using R, tidyverse, janitor, and RColorBrewer
packages. The report is built with Quarto. The code is available on
[GitHub](https://github.com/jeremy-allen/prosecution.git).